---
title: "Introduction to Data Analysis and Visualisation using R"
author: "Professor Di Cook, Econometrics and Business Statistics"
date: "Workshop for the Institute for Safety, Compensation and Recovery Research"
output:
  beamer_presentation: 
    theme: Monash
---

```{r setup, include = FALSE}
library("knitr")
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  cache = FALSE,
  fig.height = 2,
  fig.width = 5,
  fig.caption = FALSE,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library("rmarkdown")
library("devtools")
library("readr")
library("tidyr")
library("ggplot2")
library("ggthemes")
library("gridExtra")
library("dplyr")
library("lubridate")
library("GGally")
library("rworldmap")
library("ggmap")
library("scales")
library("dichromat")
library("RColorBrewer")
library("viridis")
library("purrr")
library("broom")
library("timeDate")
library("haven")
library("boot")
```

# Session 2 

**Making basic plots, grammar of graphics, good practices**

(If you re-started RStudio, be sure to re-open your project too.)

# Using the package ggplot2

Elements of a plot

- data
- aesthetics: mapping of variables to graphical elements
- geom: type of plot structure to use
- transformations: log scale, ...

Additional components 

- layers: multiple geoms, multiple data sets, annotation
- facets: show subsets in different plots
- themes: modifying style

RStudio's **[cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)** gives a nice, concise overview of the plotting capabilities.

# Data - Currency cross rates

Extracted from [http://openexchangerates.org](http://openexchangerates.org), extracted using the json api, with the R package, jsonlite.

```{r}
rates <- read_csv("data/rates.csv")
rates[1:5,1:8]
```

If you'd like to collect exchange rates yourself, see [here](https://gist.github.com/cpsievert/e05da83fc4253e6d1986).

# Plotting points

```{r}
ggplot(data=rates, aes(x=date, y=AUD)) + geom_point()
```

# Data structure

- Plots are constructed by mapping elements of data to graphical attributes. 
- Having data in a tidy structure make mapping clearer
- Some ways of making mappings make it easier for the reader to perceive structure better

# Plot structure

- data: rates 
- aesthetics: x=date, y=AUD
- geom: point, line


# Plotting lines

```{r}
ggplot(data=rates, aes(x=date, y=AUD)) + geom_line()
```

# Points and lines

```{r}
ggplot(data=rates, aes(x=date, y=AUD)) + 
  geom_line() + geom_point()
```

# Multiple currencies

```{r}
ggplot(data=rates, aes(x=date, y=AUD)) + geom_line() +
  geom_line(aes(y=NZD), colour="blue") + 
  geom_line(aes(y=GBP), colour="red")
```

# Hmmm...

- That code is clunky!
- Better to rearrange data, and then let ggplot2 handle the colors, legends, ...

# Better way

```{r}
rates.sub <- select(rates, date, AUD, NZD, GBP)
rates.sub.m <- gather(rates.sub, currency, rate, -date)
ggplot(data=rates.sub.m, aes(x=date, y=rate, colour=currency)) + geom_line() 
```

# GRAMMAR

- The grammar of graphics makes the mapping of a data variable to a plot element explicit. 
- This is a huge advance in data visualisation
- This provides a closer connection between data, plots and models.

# Mappings

- Date is mapped to position along the x axis
- Rate is mapped to position along the y axis
- Currency is mapped to colour

# Information communication

```{r}
ggplot(data=rates.sub.m, aes(x=date, y=rate, colour=currency)) + geom_line() 
```

>- What can you read from this plot? What is the main observation?
>- The cross-rates for AUD and NZD with the USD are similar, $1USD can buy approximately $1.30 of both, but the GBP is lower, and $1USD only buys 2/3 of a GBP. Do we need a plot to know this?

# Communicate trend: Scale currencies

```{r}
rates.sub <- mutate(rates.sub, AUD=scale(AUD), NZD=scale(NZD), GBP=scale(GBP))
rates.sub$date <- as.Date(rates.sub$date)
rates.sub.m <- gather(rates.sub, currency, rate, -date)
```

#

```{r}
ggplot(data=rates.sub.m, aes(x=date, y=rate, colour=currency)) +
         geom_line()
```

- Now you can read off the trend: the AUD and NZD trend similarly in this time period, but the GBP is different. The GBP goes down in cross-rate, as the AUD/NZD go up.

# Your turn

In the plot below, how are variables mapped to plot elements?

```{r echo=FALSE, fig.align='center'}
ggplot(data=rates.sub, aes(x=AUD, y=NZD, colour=date)) + geom_path() + theme(aspect.ratio=1)
```

```{r echo=FALSE, eval=FALSE}
AUD rate is mapped to position along the x axis
NZD rate is mapped to position along the y axis
first differences are mapped to lines
date is mapped to colour of line, date at beginning of line
```

# Adding marginal rug plot

```{r fig.align='center'}
ggplot(data=rates.sub, aes(x=AUD, y=NZD)) + 
  geom_point(alpha=0.2) + geom_rug(colour="red", alpha=0.3) + 
  theme(aspect.ratio=1)
```

# Other types of plots

- bar charts, pie charts
- boxplots, violins, 
- histograms
- density plots
- dotplots

Look up `?geom_histogram` and choose the index for the ggplot2 package. Look at the `geom_` options. There are many! We will only cover the few main ones.

# Type of variables suggests mapping

1. The values of **quantitative** variables should be mapped to **position along a line**, e.g. histogram, scatterplot. Mapping them to colour will yield only rough return of information to the reader. 
2. Categorical variables could be mapped to 
  - colour, if there are few categories,  
  - aggregated and mapped to position along the line, 
  - mapped to angle, if all categories are available.
3. Order is important, and if no natural order available then impose one e.g. using count

# Categorical variables - barchart

The social variables of the PISA data include internet usage. This is a subset.

```{r}
internet <- read_csv("data/internet.csv")
dim(internet)
colnames(internet)
```

# Categorical variables - barchart

```{r}
ggplot(data=internet, aes(x=`Social networks`)) + 
  geom_bar(binwidth=0.5) 
```

# Categorical variables - barchart

Simpson's paradox may be in play when there are multiple categorical variables. Need to divide it into basic elements. 

```{r}
ggplot(data=internet, aes(x=`Social networks`)) + 
  geom_bar(binwidth=0.5) +
  facet_grid(Gender~name)
```

# Categorical variables - stacked barchart

```{r}
ggplot(data=internet, aes(x=`Social networks`, fill=Gender)) + 
  geom_bar(binwidth=0.5) +
  facet_wrap(~name, ncol=5) + theme(legend.position="bottom")
```

# Categorical variables - dodged bars 

```{r}
ggplot(data=internet) + 
  geom_bar(aes(x=`Social networks`, fill=Gender),
          position="dodge") +
  facet_wrap(~name, ncol=5) + 
  theme(legend.position="bottom")
```

# Categorical variables - piechart

```{r}
ggplot(data=internet, aes(x=factor(1), fill=factor(`Social networks`))) + 
  geom_bar(width = 1) + scale_x_discrete("") +
  scale_y_continuous("") +
  scale_fill_hue("Social Network Use") +
  coord_polar(theta = "y")
```

Yes, its deliberately made hard to do !

# Quantitative and categorical - boxplots

Data are measurements from the National Research Council in the USA, evaluating graduate programs in Statistics. 

```{r}
grad <- read_csv("data/graduate-programs.csv")
dim(grad)
colnames(grad)
```

#

```{r}
ggplot(data=grad, aes(x=subject, y=AvGREs)) + 
  geom_boxplot()
```

#

```{r}
ggplot(data=grad, aes(x=subject, y=AvGREs)) + 
  geom_violin()
```

# Your turn

- Create a side-by-side boxplot of average number of publications by program
- Then answer, "how do the four programs compare in terms of average number of publications?" 

# Cognitive principles

- **Hierarchy of mappings**: (first) position along an axis - (last) color (Cleveland, 1984; Heer and Bostock, 2009)
- **Pre-attentive**: Some elements like color are noticed before you even realise it. Other elements like axes are to look up information later.
- **Color palettes**: qualitative, sequential, diverging. The type of variable determines the appropriate palette.
- **Color blindness**: you can proof your plots with te dichromat package.
- **Proximity**: To compare elements, place them close together. 
- **Change blindness**: When focus is interrupted differences may not be noticed, can occur when you are reading across multiple plots.

# Hierarchy of mappings

1. Position - common scale (BEST)
2. Position - nonaligned scale
3. Length, direction, angle
4. Area
5. Volume, curvature
6. Shading, color (WORST)

# Pre-attentive

Can you find the odd one out?

```{r echo=FALSE,}
df <- data.frame(x=runif(100), y=runif(100), cl=sample(c(rep("A", 1), rep("B", 99))))
qplot(x, y, data=df, shape=cl) + theme_bw() + theme(legend.position="None", aspect.ratio=1)
```

#

Is it easier now?

```{r echo=FALSE,}
qplot(x, y, data=df, colour=cl) + theme_bw() + theme(legend.position="None", aspect.ratio=1)
```


# Color palettes

- Qualitative: categorical variables
- Sequential: low to high numeric values
- Diverging: negative to positive values

#

```{r, echo=FALSE, fig.height=7, fig.width=12,}
display.brewer.all()
```

# Scales

```{r}
ggplot(data=internet, aes(x=`Social networks`, fill=Gender)) + 
  geom_bar(position="dodge") +
  scale_fill_manual(values=c("Female"="orange", "Male"="darkgreen")) + 
  facet_wrap(~name, ncol=5) + 
  theme(legend.position="bottom")
```

# Scales

```{r}
ggplot(data=grad, aes(x=subject, y=AvGREs)) + 
  geom_boxplot() + scale_y_log10()
```

# Axes

The date time axis is a little trickier to re-organise, but it can be done.

```{r}
rates.sub.m$date <- as.POSIXct(rates.sub.m$date)
ggplot(data=rates.sub.m, aes(x=date, y=rate, 
  colour=currency)) + geom_line() + 
  scale_x_datetime(breaks = date_breaks("1 month"), 
                   labels = date_format("%b"))
```

#

```{r}
ggplot(data=rates.sub.m, aes(x=date, y=rate, 
        colour=currency)) + geom_line() +
  xlab("Date") + ylab("Standardized rates") + 
  ggtitle("Cross rates 23/2/2015-11/11/2015")
```

```{r echo=FALSE}
p <- ggplot(data=rates.sub.m, aes(x=date, y=rate, 
  colour=currency)) + geom_line() + 
  scale_x_datetime(breaks = date_breaks("1 month"), 
                   labels = date_format("%b"))
```

# Equations in labels

```{r}
ggplot(data=rates.sub.m, aes(x=date, y=rate, colour=currency)) +
  geom_line() +
  xlab(expression(Date[i]^2~ mu ~ pi * sigma)) + 
  ylab("Standardized rates") + 
  ggtitle("Cross rates 23/2/2015-11/11/2015")
```

# Legend Position

```{r}
p + theme(legend.position = "bottom")
```


# Themes

```{r}
p + theme_tufte()
```

#

```{r}
p + theme_economist()
```

# Color palettes

```{r}
p + scale_color_brewer("", palette = "Dark2")
```

# Color blind-proofing

```{r, fig.width=4.5, fig.show='hold', fig.align='default',}
clrs <- hue_pal()(3)
p + scale_color_manual("", values=clrs) + 
  theme(legend.position = "none")
```

#

```{r, fig.width=4.5, fig.show='hold', fig.align='default',}
clrs <- dichromat(hue_pal()(3))
p + scale_color_manual("", values=clrs) + 
  theme(legend.position = "none")
```

#

```{r, fig.width=4.5, fig.show='hold', fig.align='default',}
clrs <- brewer.pal(3, "Dark2")
p + scale_color_manual("", values=clrs) + 
  theme(legend.position = "none")
```

#

```{r, fig.width=4.5, fig.show='hold', fig.align='default',}
clrs <- dichromat(brewer.pal(3, "Dark2"))
p + scale_color_manual("", values=clrs) + 
  theme(legend.position = "none")
```

# Your turn

Proximity - From with plot can you answer: Is the proportion of girls who use social networks every day (4) higher than boys, in Australia? And is this different in Germany?

```{r, echo=FALSE}
ggplot(data=internet, aes(x=`Social networks`)) + 
         geom_bar() + 
  facet_grid(Gender~name)
```


```{r echo=FALSE, fig.show='hide'}
internet.m.tb <- internet[,c(1,3,8)] %>%
                     group_by(name, Gender, `Social networks`) %>% 
                     tally(sort=TRUE) 
internet.m.tb <- subset(internet.m.tb, !is.na(`Social networks`))
internet.m.tb.n <- summarise(group_by(internet.m.tb, name, Gender), tot=sum(n)) 
internet.m.tb <- merge(internet.m.tb, internet.m.tb.n)
internet.m.tb.p <- summarise(group_by(internet.m.tb, name, Gender, `Social networks`), p=n/tot)
```

```{r, echo=FALSE, fig.show='hide'}
ggplot(data=internet.m.tb.p, aes(x=`Social networks`, y=p, color=Gender)) +
  geom_line() + 
  facet_wrap(~name, ncol=5) + theme(legend.position="bottom")
```

# Your turn

- Brainstorm with your neighbour ways to rearrange this plot to answer the question. 

- Then tackle this question: Are German girls more likely to report using social networks once or twice per month (1) than Japanese girls?

- What ways would you re-arrange the plot to tackle this one?


```{r, echo=FALSE, fig.show='hide'}
ggplot(data=internet.m.tb.p, aes(x=`Social networks`, y=p, color=name)) + 
  geom_line() + 
  facet_wrap(~Gender, ncol=2) 
```

# Proximity

- It is ok to make more than one plot. 
- Actually it is recommended. 

# Your turn

For the NY workers compensation data

- Make a barchart of the district name
- Fill the barchart by Gender, but make the height of the bars equal
- Make a violin plot of age, by claim type

```{r echo=FALSE, eval=FALSE}
workers <- read_csv(file="data/Assembled_Workers__Compensation_Claims___Beginning_2000.csv")#, n_max=100000)
dim(workers)
colnames(workers)
table(workers$`Claim Type`)
workers$`District Name` <- factor(workers$`District Name`, 
  levels=names(table(workers$`District Name`)[order(table(workers$`District Name`), decreasing=TRUE)]))
ggplot(data=workers, aes(x=`District Name`)) + geom_bar()
ggplot(data=workers, aes(x=`District Name`, fill=Gender)) + 
  geom_bar(position="fill")
ggplot(data=workers, aes(x=`Claim Type`, y=`Age at Injury`)) + 
  geom_violin()
```

# Your turn

How would you answer these questions?

- What is the most common district for injuries?
- Is the distribution of gender districts?
- Is the age of injury the same across the different claim types?

# Credits

Notes prepared by Di Cook, building on joint workshops with Carson Sievert, Heike Hofmann, Eric Hare, Hadley Wickham.
